---
title: "covid_texas_fpca_fatality"
author: "Bohan Wu"
date: "6/30/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Data preprocessing: Hospitalization Data
```{r, echo= FALSE,message=FALSE, results='hide'}
rm(list = ls()); 
```
```{r}
#library(devtools)
#install_github("drkowal/rSTAR")
library(readr)
library(tidyverse)
library(reshape2)
library(dplyr)
source("fpca_em.R")
source("james_em.R")
source("james estimator.R")
source("face estimator.R")
source("star_EM2.R")
source("estimator combined.R")
library(refund)
```
```{r}
dat_f = read_csv(file = 'Texas COVID-19 Fatality Count Data by County.csv') 
# Tidy:
dat_f = dat_f %>% pivot_longer(
  cols = colnames(dat_f)[-1], 
  names_to = "date", 
  values_to = "total fatalities") %>%
  mutate(date = as.Date(paste(date, "-2020", sep=''), "%m-%d-%Y"))

# Add the new fatalities by county:
dat_f = dat_f %>% 
  group_by(`County Name`) %>% 
  mutate(fatalities = c(NA, diff(`total fatalities`)))

county_population <- read.csv("Texas County Population.csv", row.names=1, sep=";")
# Tidy:
county_population = county_population %>%
  rename(`County Name` = County.Name)%>% 
  mutate(Population = as.numeric(Population))%>% distinct()


dat_f = dat_f %>%
  inner_join(county_population)

```
Case count data
```{r}
#----------------------------------------------------------------------------
dat_c = read_csv(file = 'Texas COVID-19 Case Count Data by County.csv') 

# Tidy:
dat_c = dat_c %>% pivot_longer(
  cols = colnames(dat_c)[-1], 
  names_to = "date", 
  values_to = "total cases") %>%
  mutate(date = as.Date(paste(date, "-2020", sep=''), "%m-%d-%Y"))

# Add the new fatalities by county:
dat_c = dat_c %>% 
  group_by(`County Name`) %>% 
  mutate(cases = c(NA, diff(`total cases`)))

# Merge cases and fatalities:
#----------------------------------------------------------------------------
dat_cf = inner_join(dat_c, dat_f)
```
###Problems: 
1. Statewide Total, NA are redudant rows. > Remove Statewide Total, NA. 
```{r visualization,tidy.opts=list(width.cutoff=60)}
#Remove Statewide Total from the dat_f
dat_cf0 = dat_cf %>% filter(`County Name` != "Total")%>% group_by(`County Name`) 
```
2. For each county, set the starting date to be the date 10 fatalities are recorded
```{r}
dat_cf.first10 = dat_cf0 %>% filter(`total cases` >= 10)%>% mutate(which_date = seq(length(fatalities))) %>% rename(county = `County Name`)
#Visualization
ggplot(dat_cf.first10, aes(x = which_date, y = fatalities, group = county )) + geom_line()
```
```{r,message=FALSE}
#Cast a matrix where columns are locations, columnas are dates
mat_f = dcast(dat_cf.first10 %>% select(c(county,which_date,fatalities)), county ~which_date,sum, fill = -1) %>% filter(complete.cases(county)) %>% column_to_rownames(var = "county") %>% as.matrix() 
mat_f[mat_f < 0] <- NA
``` 
### Remove counties with less than 10 observations
```{r}
threshold = 10
selected_counties <- apply(mat_f,1,function(row) sum(!is.na(row)) >= 10)
print(paste0(sum(selected_counties)," out of ", length(selected_counties)," are selected" ))
mat_f = mat_f[selected_counties,]

```
## FPCA 
###Set the parameters
```{r}
library(fdapace)
library(fpca)
library(gdata)
library(glmnet)
library(refund)

Lt_f <- lapply(seq(nrow(mat_f)), function(i) which(!is.na(mat_f[i,]))) #Time Grid
y_f = unmatrix(mat_f, byrow = TRUE)%>% na.exclude()  #Vectorized observations
n_counties = nrow(mat_f) #number of counties
n_dates = ncol(mat_f) #total days since the outbreak
n_fpc = 4 #number of FPCs
```
```{r}
source("face estimator.R")
source("fpca_em.R")

```
#FPCA.MLE Fitting
```{r}
#d = dat_cf.first10 %>% select(county,fatalities,which_date)
FPCA.MLE.Fit <- function(input,r.set = n_fpc){

#Create input dataset
d2 <<- melt(input)  %>% 
  rename(dates = Var2, county = Var1) %>%
  group_by(county )  %>% 
  drop_na()%>% mutate(dates = dates/n_dates)
d2 = d2[,c("county","value","dates")]

M.set<- c(r.set,r.set + 1)
ini.method="EM"
basis.method="bs"
sl.v=rep(1.5,10)
max.step= 50
  
#FPCA MLE fit
result <<- fpca.mle(as.matrix(d2),M.set, r.set,ini.method, basis.method,sl.v,max.step, grids= seq(0,1,length.out = n_dates))
##ret(mat_f)scaled grid

 grids.new<-result$grid

  M <-result$selected_model[1]
r<-result$selected_model[2]

 ## estimated eigenvalues 
 evalest<-result$eigenvalues     #

 ##estimated error variance
 sig2est<-result$error_var       

 ##estimated eigenfunctions
 eigenfest<<-result$eigenfunctions
 
# par(mfrow=c(2,3))
 #  for(i in 1:r){
 # plot(grids.new,eigenfest[i,],ylim=range(eigenfest),xlab="time",ylab=paste("eigenfunction",i))}

muest<-result$fitted_mean
fpcs<-fpca.score(d2 %>% as.matrix(),grids.new,muest,evalest,eigenfest,sig2est,r)


pred<<-fpca.pred(fpcs, muest,eigenfest) %>% t()%>% melt()
ggplot(pred) + geom_line(aes(x = Var2,y = value, group = Var1))

opt  = list(fitted.values = matrix(), coefficients = matrix())
opt$fitted_values = fpca.pred(fpcs, muest,eigenfest) %>% t()
opt$coefficients = c(as.vector(fpcs),muest, as.vector(eigenfest))
return(opt)
}


```
###Estimator 1: Construct Estimator Function Using fpca.face
```{r}
estimator.mle = function(v = y_f)
{
input <<- list(Ly = matrix(nrow = n_counties, ncol = n_dates),Lt = Lt_f,all.Ly = v)

for(i in seq(n_counties)){
  input$Ly[i,input$Lt[[i]]] <- input$all.Ly[input$Lt[[i]]]
  input$all.Ly <- input$all.Ly[-input$Lt[[i]]]
}
#Use first 4 FPCs

FPCA.fit <- FPCA.MLE.Fit(input$Ly,n_fpc )

fitted_f.obs = FPCA.fit$fitted_values
#Output
output <- list(fitted.values = matrix(), coefficients = matrix())

for(i in seq(n_counties)){
  output$fitted.values <- append(output$fitted.values,fitted_f.obs[i, input$Lt[[i]]])
}
output$coefficients <- FPCA.fit$coefficients%>% na.exclude()
output$fitted.values <- output$fitted.values  %>% na.exclude()
return(output)
}
d_1 = estimator.mle()
```


##RSTAR Model Fit: FPCA + STAR EM
```{r freq-lm,message=FALSE,tidy.opts=list(width.cutoff=60)}
library(rSTAR)
library(gt)

#transformation = 'log' # Log transformation
#transformation = 'np' 
# EM algorithm for STAR (using the log-link)
fit_em = star_EM(y = y_f, 
                 estimator = estimator.fpca, 
                 transformation = 'log',
  sd_init = 5,
  tol = 10^-6,
  max_iters = 1000)


```

##MLE 
###Plot the progression of log-likelihood over time: log_transform
```{r}
plot(fit_em$logLik_all, main = "log-likehood under log tranformation")
which.max(fit_em$logLik_all)
```
```{r}
y_hat = fitted(fit_em)
mat_f_hat = mat_f
for(i in seq(n_counties)){
  mat_f_hat[i,input$Lt[[i]]] <- y_hat[input$Lt[[i]]]
  y_hat <- y_hat[-input$Lt[[i]]]
}
```
```{r}
dat_f_est = melt(mat_f_hat) %>% rename('county'  = Var1,which_date = Var2,fatalities_est = value)
dat_cf_all = dat_cf.first10 %>% inner_join(dat_f_est,by = c("county","which_date"))
#Visualization
p1 = ggplot(dat_cf_all, aes(x = which_date, y = fatalities, group = county)) + geom_line() + geom_line(aes(x = which_date, y = fatalities_est, group = county ,color = "estimated"))+
  ggtitle("Actual vs Estimated Fatalities (W/o James EM)")
```
#Eigenfunctions
```{r}
theta_EM1 = fit_em$coefficients
ef1 = matrix(theta_EM1[(length(theta_EM1) - n_fpc - n_fpc*n_dates):(length(theta_EM1) - n_fpc - 1)], nrow = n_dates )
dat_phis = t(ef1) %>% melt() %>% rename(which_date = Var2, which_fpc = Var1) %>% mutate(which_fpc = as.factor(which_fpc))
ef_plot1 = ggplot(dat_phis) + aes(x = which_date, y = value, color = which_fpc,group = which_fpc) +
  geom_line() +
  ggtitle("FPCs (W/o James EM)")+ scale_fill_continuous(rainbow(5))
```
###Plot the progression of log-likelihood over time: nonparametric transform
```{r}
# EM algorithm for STAR (using the log-link)
fit_em.np = star_EM(y = y_f, 
                 estimator = estimator.fpca, 
                 transformation = 'np' ,
  sd_init = 10,
  tol = 10^-10,
  max_iters = 100)
plot(fit_em, main = "log-likehood under nonparametric tranformation")

```

##RSTAR Model Fit: EM^2 framework
```{r}
fit_sfpca_em = STAR_FPCA_EM(y = y_f, 
             estimator_em = james_estimator,
             estimator_initialize = estimator.fpca, 
                 transformation = "log",
  sd_init = 5,
  tol = 10^-6,
  max_iters = 100)
# 
# fit_em2 = star_EM(y = y_f, 
#                  estimator = estimator.combined, 
#                  transformation = 'log' ,
#   sd_init = 10,
#   tol = 10^-10,
#   max_iters = 100)
```
Loglikelihood Check
```{r}
plot(fit_sfpca_em$logLik_all, main = "log-likehood with EM^2")
which.max(fit_sfpca_em$logLik_all)
length(fit_sfpca_em$logLik_all)
```

###Fitted values:
```{r}
y_hat = fitted(fit_sfpca_em)
mat_f_hat = mat_f
for(i in seq(n_counties)){
  mat_f_hat[i,input$Lt[[i]]] <- y_hat[input$Lt[[i]]]
  y_hat <- y_hat[-input$Lt[[i]]]
}
```
```{r}
dat_f_est = melt(mat_f_hat) %>% rename('county'  = Var1,which_date = Var2,fatalities_est = value)
dat_cf_all = dat_cf.first10 %>% inner_join(dat_f_est,by = c("county","which_date"))
#Visualization
p2 = ggplot(dat_cf_all, aes(x = which_date, y = fatalities, group = county)) + geom_line() + geom_line(aes(x = which_date, y = fatalities_est, group = county ,color = "estimated"))+
  ggtitle("Actual vs Estimated (W/ James EM)")
```
#Eigenfunctions
```{r}
theta_EM2 = fit_sfpca_em$coefficients
ef = matrix(theta_EM2[(length(theta_EM2) - n_fpc - n_fpc*n_dates):(length(theta_EM2) - n_fpc - 1)], nrow = n_dates )
dat_phis = t(ef) %>% melt() %>% rename(which_date = Var2, which_fpc = Var1) %>% mutate(which_fpc = as.factor(which_fpc))
ef_plot2 = ggplot(dat_phis) + aes(x = which_date, y = value, color = which_fpc,group = which_fpc) +
  geom_line() +
  ggtitle("FPCs (W/ James EM)")+ scale_fill_continuous(rainbow(5))
```
```{r}
library(gridExtra)
grid.arrange(p1, p2,ef_plot1, ef_plot2,nrow = 2, ncol=2)

par(mfrow = c(1,2))
plot(fit_em$logLik_all, main = "log-likehood with Face Estimator")
plot(fit_sfpca_em$logLik_all, main = "log-likehood with EM^2")
abline(h = max(fit_em$logLik_all), col = 'coral2', lwd = 2)


```
